---
title: "Bootcamp 4: Chloropleth maps"
author: "Lucien Baumgartner"
date: "10/26/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=F)
```

```{css echo=F}
pre code, pre, code {
  white-space: pre !important;
  overflow-x: scroll !important;
  word-break: keep-all !important;
  word-wrap: initial !important;
}

pre.html{
  background-color:white;
}

pre.r{
  background-color:black;
  color:white;
}
```


Okay, let's go down to business. In this bootcamp, we will produce a chloroppleth map. The starting point is the following question:

__Which districts have a lot of households with children?__

I will first paraphrase this question to:

__What is the median number of children (per household) per district over the last 5 years?__

We will see that the answer to this question is simple: zero. Why? Singles kinda rule I guess.

Hence, I rephrase the whole thing to:

__What is the median share of households with children per district over the last 5 years?__

This should serve a reminder: even if your initial idea is stupid, think about it differently (think about *maximizing variance*), and try again.

The following script includes my whole workflow, including some exploratory stuff, just so that you see how I'd go about.

```{r fig.align='center', fig.width=12, fig.height=12}
library(dplyr) # data crunching
library(ggplot2) # plots
library(rgdal) # geodata handling
library(viridis) # color scale for lazy people
library(ggrepel) # text label repel
library(extrafont) # different fonts

# clean workspace
rm(list=ls())

# all the registered fonts
fonts()

# load custom functions available at: https://github.com/lucienbaumgartner/r-helpers
source('~/r-helpers/ggplot/ggplot-helper.R')

# set WD
setwd('~/ddj18/output/')

# load pop data
load('01-bevoelkerung-clean.RData')
str(df)

# load shapefile
shp.raw <- readOGR('~/ddj18/input/st-zh/shapefile/Quartier_Shapefile/', layer = 'StatistischesQuartier')

# inspect shp meta info
str(shp.raw@data)
# we have both kreis and distritict vars!
# this means we will have to join the data by one of the two

## let's compute the median number of children over the last 5 years per district
# number of obs per district
table(df$quarlang)
# number of obs per district in the last five years
df %>% 
  filter(stichtagdatjahr%in%2012:2017) %>% 
  group_by(quarlang) %>% 
  summarise(n=n()) %>% 
  print(n=100)

# actual stuff that we are interested in:
df %>% 
  filter(stichtagdatjahr%in%2012:2017) %>% 
  summarise(n.kids=median(anzahlkinder)) %>% 
  print(n=100)
# ... mmh okay.. that's kinda stupid to plot...
# let's compute the share of household with more than 1 kid per district over the last 5 years instead...
kids <- df %>% mutate(has.kids=ifelse(anzahlkinder>0,1,0)) %>% 
  filter(stichtagdatjahr%in%2012:2017) %>% 
  group_by(quarlang, stichtagdatjahr, has.kids) %>% 
  summarise(n=n()) %>%
  mutate(p=n/sum(n)) %>% 
  ungroup %>% 
  group_by(quarlang, has.kids) %>% 
  summarise(median.p=median(p)) %>% 
  print(n=100)
# already looks more promising

# now we have to check whether the names of the districts in shp actually match those in the data!!!
table(unique(shp.raw$QNAME)%in%unique(kids$quarlang))
# yep! this makes everything a lot easier!

## now let's join the data to the shp
# since we work with ggplot, and we don't do super fancy stuff with geodata, we just coerce the shp to a dataframe
class(shp.raw)
shp <- fortify(shp.raw, region='QNAME') # use QNAME as region (otherwise it will be dropped)
class(shp)
# dims before join
dim(shp)
# join; NOTE: use a left join!!!
# shp$region is now shp$id; to join we will have to rename kids$quarlang to kids$id
shp <- left_join(shp, rename(kids, id=quarlang))
# dims after join
dim(shp)
# the number of rows doubled and we got three additional variables...
# why? because kids$quarlang contains every district twice (because we grouped by quarlang AND has.kids, and the latter has 2 distinct values (0,1))
# the three addtional variables are all other var in kids (those are the reason why we actually joined the data...)
# since we will only plot the share of the pop who has children, we can drop all the others and directly proceed to the plotting
kp <- shp %>% 
  filter(has.kids==1) %>% 
  ggplot(.) + # feed the subset to the ggplot
  geom_polygon(aes(x=long, y=lat, group=group, fill=median.p),
               color='white',
               size=0.2) + # plot the polygon and fill it with the percentages
  coord_equal() + # make x and y equistant
  scale_fill_viridis(option = 'D', name='Percentage',  label=scales::percent) +
  labs(title='Median share of households with kids over the last 5 years per district',
    subtitle='Some additional blahblah explaining some very interesting stuff'
    ) +
  # get rid of the background, axis, grids, etc, and add title and subtitle spacing
  theme_empty(
    title.spacing = 10,
    subtitle.spacing = 10
  ) +
  # add another background, change text color
  theme(
    plot.background = element_rect(fill='black'),
    legend.background = element_rect(fill='black'),
    text = element_text(colour='white', family = 'Source Code Pro'),
    plot.title = element_text(size=17),
    plot.subtitle = element_text(size=15),
    legend.title = element_text(size=15),
    legend.text = element_text(size=12)
  )
kp
  
# now we would like to show the reader in which people with kids add up to 20% of the district pop
# for that we need to compute the anker of the annotation: in this case the district polgyon centroids
centroids <- getSpPPolygonsLabptSlots(shp.raw)
centroids <- centroids[order(shp.raw@data$QNAME),]

# join points to the shp via QNAME
kids <- left_join(kids,
          cbind(as_tibble(centroids), 
                as.character(shp.raw@data$QNAME)[order(shp.raw@data$QNAME)]) %>% 
            setNames(., c('long.c', 'lat.c', 'quarlang')),
          by='quarlang')
# add everything to the plot
kp <- kp +
  geom_point(data=filter(kids, has.kids==1&median.p>.2), 
             aes(x=long.c, y=lat.c),
             color='white') +
  geom_text_repel(data=filter(kids, has.kids==1&quarlang%in%c('Friesenberg', 'Leimbach')), 
            aes(x=long.c, y=lat.c, label=paste0(quarlang,
                                                '\n',
                                                format(round(median.p*100,1), digits = 1),
                                                '%')
                ),
            nudge_y = -1000, 
            nudge_x = -2000, 
            size=5, 
            segment.size = 0.2, 
            color='white', 
            family='Source Code Pro') +
  geom_text_repel(data=filter(kids, has.kids==1&quarlang%in%c('Saatlen')), 
                  aes(x=long.c, y=lat.c, label=paste0(quarlang,
                                                      '\n',
                                                      format(round(median.p*100,1), digits = 1),
                                                      '%')
                      ),
                  nudge_y = 1000, 
                  nudge_x = 2000, 
                  size=5, 
                  segment.size = 0.2, 
                  color='white', 
                  family='Source Code Pro')
kp
```

