---
title: 'Bootcamp 7: Extend discrete color palettes'
author: "Lucien Baumgartner"
date: "11/6/2018"
output: 
  html_document:
    toc: true
    number_sections: true
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=F)
```

```{css echo=F}
pre code, pre, code {
  white-space: pre !important;
  overflow-x: scroll !important;
  word-break: keep-all !important;
  word-wrap: initial !important;
}

pre.html{
  background-color:white;
}

pre.r{
  background-color:black;
  color:white;
}
```

# Problem
```{r}
library(gplots)
library(ggplot2)
library(dplyr)
library(RColorBrewer)
library(viridis)
library(extrafont) # different fonts

rm(list=ls())

source('~/r-helpers/ggplot/ggplot-helper.R')

# set WD
setwd('~/ddj18/output/')

# load pop data
load('01-bevoelkerung-clean.RData')
```


If we plot the data without extending the color palette, we end up with the following result:
```{r fig.align='center', fig.width=12, fig.height=8}
# aggregate data
kids <- df %>% mutate(has.kids=ifelse(anzahlkinder>0,1,0)) %>% 
  filter(stichtagdatjahr%in%2012:2017) %>% 
  group_by(quarlang, stichtagdatjahr, has.kids) %>% 
  summarise(n=n()) %>%
  mutate(p=n/sum(n)) %>% 
  ungroup %>% 
  group_by(quarlang, has.kids) %>% 
  summarise(median.p=median(p))

# simple bar plot
kids %>% 
  filter(has.kids==1) %>% 
  ggplot() + 
  geom_bar(aes(quarlang, median.p, fill=factor(quarlang)), stat='identity') +
  scale_fill_brewer(palette="Set2") +
  theme(axis.text.x = element_text(angle=45, hjust=1))
```

# Brewer palettes
Now, to extend the color palette of your choice, you don't need anything else than the palette (dah), and the `colorRampPalette()` function native to the graphic devices.

```{r fig.align='center', fig.width=12, fig.height=8}
# brewer palettes
display.brewer.all()

# define the number of discrete colors you need (== factor levels of the fill or colour var)
c.count <- length(unique(kids$quarlang))

# feed the palette to the function to create a pre-parametrized function
get.palette <- colorRampPalette(brewer.pal(9, "Set1"))

# function
get.palette

# function fed with the umber of colours needed
get.palette(c.count)

# is everything of the same length?
length(get.palette(c.count))==c.count&c.count==length(unique(kids$quarlang))

kids %>% 
  filter(has.kids==1) %>% 
  ggplot() + 
  geom_bar(aes(quarlang, median.p, fill=factor(quarlang)), stat='identity') + 
  theme(legend.position="right") +
  scale_fill_manual(values=get.palette(c.count))  +
  theme(axis.text.x = element_text(angle=45, hjust=1))
```

# Custom colors: via expanding
Another example with custom colors:
```{r fig.align='center', fig.width=12, fig.height=8}
# draw some random colors
set.seed(727678)
c <- sample(colors(), 7) %>% col2hex
c

get.palette <- colorRampPalette(c)

kids %>% 
  filter(has.kids==1) %>% 
  ggplot() + 
  geom_bar(aes(quarlang, median.p, fill=factor(quarlang)), stat='identity') + 
  # HERE you specify the colors:
  scale_fill_manual(values=get.palette(c.count))  +
  theme(axis.text.x = element_text(angle=45, hjust=1))
```

# Custom colors: direct approach
Alternative example with custom colors:
```{r fig.align='center', fig.width=12, fig.height=8}
# draw some random colors
set.seed(727678)
c <- sample(colors(), c.count) %>% col2hex
c

kids %>% 
  filter(has.kids==1) %>% 
  ggplot() + 
  geom_bar(aes(quarlang, median.p, fill=factor(quarlang)), stat='identity') + 
  scale_fill_manual(values=c) +
  theme(axis.text.x = element_text(angle=45, hjust=1))
```

# Viridis
Alternative example with viridis:
```{r fig.align='center', fig.width=12, fig.height=8}
kids %>% 
  filter(has.kids==1) %>% 
  ggplot() + 
  geom_bar(aes(quarlang, median.p, fill=factor(quarlang)), stat='identity') + 
  # HERE you specify discreteness
  scale_fill_viridis(discrete = T) +
  theme(axis.text.x = element_text(angle=45, hjust=1))
```

